---
title: "Age and Sex signature analysis"
author: "MSoria"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
  html_notebook:
    theme: lumen
    toc_float: true
    toc: yes
---

```{r load library, include=FALSE}
# Load Libraries 
# Installation of some useful packages
suppressPackageStartupMessages({c(
  library(tidyverse), 
  library(DESeq2), 
  library(dplyr), 
  library(RColorBrewer), 
  library(circlize),
  library(ggplot2), 
  library(ggrepel), 
  library(PCAtools),
  library(ggpubr),
  library(viridis),
  library(mitch),
  library(msigdbr),
  library(biomaRt),
  library(gplots)
  )})

# Load custom functions 
source("functions.R") 

set.seed(42)
```

# Load and transform data
Load counts and biological data
```{r}
cts <- readRDS("rds/cts.rds")

pheno_data <- readRDS("rds/pheno_data.rds")

# make sure colnames and rownames of the dataframes match
all(rownames(pheno_data) == colnames(cts)) # should return TRUE
```

Load DE results and datasets
```{r}
deRes_youngVSold_male_cells <- readRDS("rds/deRes_youngVSold_male_cells.rds")
deRes_youngVSold_male_tissue <- readRDS("rds/deRes_youngVSold_male_tissue.rds")

# Get gene names from biomaRt; use mirrors if default does not work
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
               # host = "https://useast.ensembl.org/",
               # host = "https://asia.ensembl.org/")

biomart_transcript2gene <- getBM(attributes = c('ensembl_gene_id',
                            'external_gene_name'), mart = mart)
```

# Supplementary Fig 5: Age signature (All)
Create fold-changes matrices and plots
```{r }
cellFC_05 <- diff_expr2(deRes_youngVSold_male_cells[, -7], 0.05, 1)
tissueFC_05 <- diff_expr2(deRes_youngVSold_male_tissue[, -7], 0.05, 1)

cellFC <- cellFC_05 %>% 
  rownames_to_column(var = "ensembl_gene") %>% 
  dplyr::select(c(ensembl_gene, log2FoldChange, padj, diffexpressed)) 
names(cellFC) <- c("ensembl_gene", "L2FC_Cell", "padj_Cell", "diff_Cell")

tissueFC <- tissueFC_05 %>% 
  rownames_to_column(var = "ensembl_gene") %>% 
  dplyr::select(c(ensembl_gene, log2FoldChange, padj, diffexpressed)) 
names(tissueFC) <- c("ensembl_gene", "L2FC_Tissue", "padj_Tissue", "diff_Tissue")

FC_mat_all <- inner_join(cellFC, tissueFC) %>% 
  mutate(Expression = ifelse(diff_Cell == "NO" & diff_Tissue == "NO", "NS",
                    ifelse(diff_Cell == "NO" & diff_Tissue == "UP", "Tissue (+)",
                    ifelse(diff_Cell == "NO" & diff_Tissue == "DOWN", "Tissue (-)",
                    ifelse(diff_Cell == "UP" & diff_Tissue == "NO", "Cell (+)", 
                    ifelse(diff_Cell == "UP" & diff_Tissue == "UP", "All (+)",
                    ifelse(diff_Cell == "UP" & diff_Tissue == "DOWN", "Cell (+) : Tissue (-)", 
                    ifelse(diff_Cell == "DOWN" & diff_Tissue == "NO", "Cell (-)", 
                    ifelse(diff_Cell == "DOWN" & diff_Tissue == "UP", "Cell (-) : Tissue (+)",
                    ifelse(diff_Cell == "DOWN" & diff_Tissue == "DOWN", "All (-)",
                                  "NA")))))))))) 
                           
biomart_transcript2gene <- biomart_transcript2gene %>% 
  rename("ensembl_gene_id" = "ensembl_gene") %>% unique()

FC_mat_all <- left_join(FC_mat_all, biomart_transcript2gene)


b <- ggplot(FC_mat_all, aes(x=L2FC_Cell, y=L2FC_Tissue, color=Expression))+ 
                       #label=external_gene_name)) + 
  geom_point(size = 4, alpha=0.8) + 
  # ggrepel::geom_label_repel(data=subset(FC_mat_all, Expression == "All (+)" | 
  #                         Expression == "All (-)"), force = 3, seed = 42,
  #                         label.padding = 0.1, point.padding = 0.2, size = 2, color = "brown") +
  labs(x = "Cell Log2FC", y = "Tissue Log2FC") +
  scale_color_viridis(discrete=TRUE, option="turbo") +
  theme_pubr() +
  geom_hline(yintercept=0, linetype="dashed", color = "lightgrey") +
  geom_vline(xintercept=0, linetype="dashed", color = "lightgrey")

b
```

# Fig 5A: Age signature (Significant data points only)
Padj < 0.05, log2FC > 1, < -1
```{r}
FC_mat_05 <- FC_mat_all %>% 
  filter(Expression !="NS")

c <- ggplot(FC_mat_05, aes(x=L2FC_Cell, y=L2FC_Tissue, color=Expression, 
                       label=external_gene_name)) + 
  geom_point(size = 4, alpha=0.8) + 
  ggrepel::geom_label_repel(data=subset(FC_mat_all, Expression == "All (+)" |
                          Expression == "All (-)"), force = 3, seed = 42,
                          label.padding = 0.1, point.padding = 0.2, size = 2, color = "brown") +
  labs(x = "Cell Log2FC", y = "Tissue Log2FC") +
  scale_color_viridis(discrete=TRUE, option="turbo") +
  theme_pubr() +
  geom_hline(yintercept=0, linetype="dashed", color = "lightgrey") +
  geom_vline(xintercept=0, linetype="dashed", color = "lightgrey")

c
```
```{r eval=FALSE, include=FALSE}
pdf("figures/ageSigx_cellVStissue_log2fc_padj05.pdf",
    width = 9,
    height = 5)
b
c
dev.off()

base_size = 15
```

# Supplementary Fig 5B: Sex Signature (All)
Load DE results 
```{r}
deRes_maleVSfemale_young_cell <- readRDS("rds/deRes_maleVSfemale_young_cell.rds")
deRes_maleVSfemale_young_tissue <- readRDS("rds/deRes_maleVSfemale_young_tissue.rds")
```

Create fold-changes matrices and plots
```{r fig.height=5, fig.width=9}
cellFC_05 <- diff_expr2(deRes_maleVSfemale_young_cell[, -7], 0.05, 1)
tissueFC_05 <- diff_expr2(deRes_maleVSfemale_young_tissue[, -7], 0.05, 1)
 
cellFC <- cellFC_05 %>% 
  rownames_to_column(var = "ensembl_gene") %>% 
  dplyr::select(c(ensembl_gene, log2FoldChange, padj, diffexpressed)) 
names(cellFC) <- c("ensembl_gene", "L2FC_Cell", "padj_Cell", "diff_Cell")

tissueFC <- tissueFC_05 %>% 
  rownames_to_column(var = "ensembl_gene") %>% 
  dplyr::select(c(ensembl_gene, log2FoldChange, padj, diffexpressed)) 
names(tissueFC) <- c("ensembl_gene", "L2FC_Tissue", "padj_Tissue", "diff_Tissue")

FC_mat_all <- inner_join(cellFC, tissueFC) %>% 
  mutate(Expression = ifelse(diff_Cell == "NO" & diff_Tissue == "NO", "NS",
                    ifelse(diff_Cell == "NO" & diff_Tissue == "UP", "Tissue (+)",
                    ifelse(diff_Cell == "NO" & diff_Tissue == "DOWN", "Tissue (-)",
                    ifelse(diff_Cell == "UP" & diff_Tissue == "NO", "Cell (+)", 
                    ifelse(diff_Cell == "UP" & diff_Tissue == "UP", "All (+)",
                    ifelse(diff_Cell == "UP" & diff_Tissue == "DOWN", "Cell (+) : Tissue (-)", 
                    ifelse(diff_Cell == "DOWN" & diff_Tissue == "NO", "Cell (-)", 
                    ifelse(diff_Cell == "DOWN" & diff_Tissue == "UP", "Cell (-) : Tissue (+)",
                    ifelse(diff_Cell == "DOWN" & diff_Tissue == "DOWN", "All (-)",
                                  "NA")))))))))) 

FC_mat_all <- left_join(FC_mat_all, biomart_transcript2gene)


a <- ggplot(FC_mat_all, aes(x=L2FC_Cell, y=L2FC_Tissue, color=Expression, 
                       label=external_gene_name)) + 
  geom_point(size = 4, alpha=0.8) + 
  # ggrepel::geom_label_repel(data=subset(FC_mat_all, Expression == "All (+)" | 
  #                         Expression == "All (-)"), force = 3, seed = 42,
  #                         label.padding = 0.1, point.padding = 0.2, size = 4, color = "brown") +
  labs(x = "Cell Log2FC", y = "Tissue Log2FC") +
  scale_color_viridis(discrete=TRUE, option="turbo") + 
  theme_pubr(base_size = 15) +
  geom_hline(yintercept=0, linetype="dashed", color = "lightgrey") +
  geom_vline(xintercept=0, linetype="dashed", color = "lightgrey")

a
```

# Fig 6A: Sex Signature ((Significant data points only)
Padj < 0.05, log2FC > 1, < -1
```{r fig.height=5, fig.width=9}
FC_mat_05 <- FC_mat_all %>% 
  filter(Expression !="NS")

b <- ggplot(FC_mat_05, aes(x=L2FC_Cell, y=L2FC_Tissue, color=Expression, 
                       label=external_gene_name)) + 
  geom_point(size = 4, alpha=0.8) + 
  ggrepel::geom_label_repel(data=subset(FC_mat_all, Expression == "All (+)" | 
                          Expression == "All (-)"), force = 3, seed = 42,
                          label.padding = 0.1, point.padding = 0.2, size = 4, color = "brown") +
  labs(x = "Cell Log2FC", y = "Tissue Log2FC") +
  scale_color_viridis(discrete=TRUE, option="turbo") +
  theme_pubr(base_size = 15) +
  geom_hline(yintercept=0, linetype="dashed", color = "lightgrey") +
  geom_vline(xintercept=0, linetype="dashed", color = "lightgrey")

b
```

```{r eval=FALSE, include=FALSE}
pdf("figures/seSigx_cellVStissue_log2fc_padj05.pdf",
    width = 9,
    height = 5)
a
b
dev.off()

```

# Figure 5B and 5C: Sex Signature (mitch)
```{r}

reactome <- msigdbr(species = "Homo sapiens", 
                 category = "C2", subcategory = "CP:REACTOME") %>% 
  dplyr::distinct(gs_name, human_ensembl_gene) %>% 
  mutate(gs_name = gsub("REACTOME_", "", gs_name)) %>% 
  mutate(gs_name = gsub("_", " ", gs_name)) %>% 
  group_by(gs_name) %>%
  summarise(all.genes = list(unique(human_ensembl_gene))) %>%
  deframe()

  df1 <- deRes_youngVSold_male_cells[, -7] %>%
    rownames_to_column(., var = "ensembl_gene_id") %>%
    group_by(ensembl_gene_id) %>% 
    dplyr::summarise(across(everything(), mean))
  
  df2 <- deRes_youngVSold_male_tissue[, -7] %>%
    rownames_to_column(., var = "ensembl_gene_id") %>%
    group_by(ensembl_gene_id) %>% 
    dplyr::summarise(across(everything(), mean))
  
mitch_reslist <- list("Cell" = df1, "Tissue" = df2)

y<-mitch_import(mitch_reslist,"DESeq2", geneIDcol="ensembl_gene_id")
head(y)

# prioritisation by effect
res<-mitch_calc(y,reactome,priority="effect")
mitch_report(res,"ageSignature.html")
#mitch_plots(res,"ageSignature.pdf")

```

```{r fig.height=12, fig.width=9.5}
d <- ncol(res$input_profile)
resrows <- length(res$detailed_sets)

hmapx <- head(res$enrichment_result[, 4:(4 + d - 1)], resrows)
rownames(hmapx) <- head(res$enrichment_result$set, resrows)
colnames(hmapx) <- gsub("^s.", "", colnames(hmapx))

my_palette <- colorRampPalette(c("blue", "white", "red"))(n = 25)
        
heatmap.2(as.matrix(hmapx), scale = "none", margins = c(5, 35),
                  dendrogram = "row", cexRow = 0.8, trace = "none", 
          cexCol = 1.2, col = my_palette, key =F )
```


```{r fig.height=12, fig.width=9.5, include=FALSE}
pdf("figures/ageSignature_mitch_hmap.pdf",
    width = 9.5,
    height = 12)

heatmap.2(as.matrix(hmapx), scale = "none", margins = c(5, 35),
                  dendrogram = "row", cexRow = 0.8, trace = "none", 
          cexCol = 1.2, col = my_palette, key =F )
dev.off()
```


# Figure 6B and 6C: Sex Signature (mitch)
```{r}

  df3 <- deRes_maleVSfemale_young_cell[, -7] %>%
    rownames_to_column(., var = "ensembl_gene_id") %>%
    group_by(ensembl_gene_id) %>% 
    dplyr::summarise(across(everything(), mean))
  
  df4 <- deRes_maleVSfemale_young_tissue[, -7] %>%
    rownames_to_column(., var = "ensembl_gene_id") %>%
    group_by(ensembl_gene_id) %>% 
    dplyr::summarise(across(everything(), mean))
  
mitch_reslist <- list("Cell" = df3, "Tissue" = df4)

y<-mitch_import(mitch_reslist,"DESeq2", geneIDcol="ensembl_gene_id")
head(y)

# prioritisation by effect
res_sex<-mitch_calc(y,reactome,priority="effect",cores=2)
mitch_report(res_sex,"sexSignature.html")
#mitch_plots(res_sex,"sexSignature.pdf")

```


```{r fig.height=12, fig.width=12}
d <- ncol(res_sex$input_profile)
resrows <- length(res_sex$detailed_sets)

hmapx <- head(res_sex$enrichment_result[, 4:(4 + d - 1)], resrows)
rownames(hmapx) <- head(res_sex$enrichment_result$set, resrows)
colnames(hmapx) <- gsub("^s.", "", colnames(hmapx))

my_palette <- colorRampPalette(c("blue", "white", "red"))(n = 25)
        
heatmap.2(as.matrix(hmapx), scale = "none", margins = c(5, 45),
                  dendrogram = "row", cexRow = 0.8, trace = "none", 
          cexCol = 1.2, col = my_palette, key =F )
```


```{r fig.height=12, fig.width=12}
pdf("figures/sexSignature_mitch_hmap.pdf",
    width = 12,
    height = 12)

heatmap.2(as.matrix(hmapx), scale = "none", margins = c(5, 45),
                  dendrogram = "row", cexRow = 0.8, trace = "none", 
          cexCol = 1.2, col = my_palette, key =F )
dev.off()
```


# Supplementary Figure 6
These figures are insides the generated mitch reports


```{r}
sessionInfo()  
```