---
title: 'Incorporating differentiation scores to Cell Study'
author: "MSoria"
date: "2024-01-18"
output:
  html_document:
    toc: yes
  html_notebook:
    theme: lumen
    toc: yes
---

Here, a differentiation "ruler" is built to be applied to the human cell and tissue data. The differentiation scores will be derived from the GSE168897 dataset.

```{r load library, include=FALSE}
# Load Libraries

# Installation of some useful packages
suppressPackageStartupMessages({c(
  library(tidyverse),
  library(DESeq2),
  library(dplyr), 
  library(msigdbr),
  library(ggplot2)
  )})

set.seed(42)
```

# Load and transform data
Load counts and biological data for cell differentiation
```{r}

url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE168897&format=file&file=GSE168897%5Fgenewise%2Dcount%2Ecsv%2Egz"
destfile <- "data/GSE168897_genewise-count.csv.gz"

# Download the file
if(!file.exists(destfile))(download.file(url, destfile))

# Read the gzipped CSV file
cts <- read.csv(gzfile(destfile)) %>% 
  column_to_rownames(var = "gene") %>% 
  as.data.frame()

pheno_data <- data.frame(ColName = names(cts)) %>% 
  separate_wider_delim(ColName, delim = "_", 
                       names = c("Cell_line", "TimePoint", "Replicate"),
                       cols_remove = F) %>% 
  mutate(ColName = paste0(Cell_line, "_", TimePoint)) %>% 
  as.data.frame()
rownames(pheno_data) <- names(cts)

# Set factors
pheno_data$TimePoint <- as.factor(pheno_data$TimePoint)
pheno_data$ColName <- as.factor(pheno_data$ColName)

# check rownames of sample data equal column names of count matrix
all(rownames(pheno_data) == colnames(cts))

# Create a deseq object with count and bio data. 
dds0 <- DESeqDataSetFromMatrix(countData = cts, colData = pheno_data, design = ~ TimePoint)

# Collapse R1 and R2 as replicates
#ddsColl <- collapseReplicates(dds0, dds0$ColName, dds0$Replicate)
ddsColl <- collapseReplicates(dds0, dds0$ColName, dds0$Replicate)

# check that the sum of the counts for "sample1" is the same
# as the counts in the "sample1" column in ddsColl
matchFirstLevel <- dds0$ColName == levels(dds0$ColName)[1]
all(rowSums(counts(dds0[,matchFirstLevel])) == counts(ddsColl[,1])) #Should return true

cts_diff <- as.data.frame(counts(ddsColl))
pheno_diff <- as.data.frame(ddsColl@colData)

head(cts_diff)

``` 

Load counts data for cell samples
```{r}
cts <- readRDS("rds/cts.rds")

pheno_data <- readRDS("rds/pheno_data.rds") %>% 
  mutate(Phenotype = paste0(Sex, "-", Age)) 

# make sure colnames and rownames of the dataframes match
all(rownames(pheno_data) == colnames(cts)) # should return TRUE

pheno_cells <- pheno_data %>% 
  filter(Type == "Cells")

cts_cells <- cts[,colnames(cts) %in% pheno_cells$RowName]

head(cts_cells)
```


***
# Supplementary Fig 4: Weighted sum using PCA for differentiation scale
Calculate differentiation scale
```{r fig.height=5, fig.width=8}
dds <- DESeqDataSetFromMatrix(countData = cts_diff, colData = pheno_diff, design = ~ TimePoint)

x <- dim(dds)

keep <- rowMeans(counts(dds)) >= 10
dds <- dds[keep,]

dds <- DESeq(dds)
vsd <- vst(dds, blind = T)
normalized_counts <- as.matrix(assay(vsd))

# Transpose the normalized count matrix (genes as columns, samples as rows) for PCA
pca <- prcomp(t(normalized_counts), scale. = TRUE)

plotPCA(vsd, intgroup="TimePoint")

# Extract the loadings for the first principal component (PC1)
pc1_loadings <- pca$rotation[, 1]

normalized_counts <- normalized_counts[names(pc1_loadings), ]
identical(names(pc1_loadings), rownames(normalized_counts))

# Compute weighted sum using PC1 loadings as weights
weighted_sum_pca <- apply(normalized_counts, 2, function(sample) {
  sum(sample * pc1_loadings, na.rm = TRUE)
})


# Min-max scaling
min_score <- min(weighted_sum_pca)
max_score <- max(weighted_sum_pca)
differentiation_score_pca <- (weighted_sum_pca - min_score) / (max_score - min_score)
differentiation_score_DF <- as.data.frame(differentiation_score_pca) %>% 
  rownames_to_column(var = "TimePoint") %>% 
  mutate(TimePoint = gsub(".*_", "", .$TimePoint)) %>% 
  group_by(TimePoint) %>% 
  summarise_all(., mean)

differentiation_score_DF

write.csv(differentiation_score_DF, "~/RA/fourCoreGenotype/full_data/data/differentiation_score_DF.csv")

a <- ggplot(differentiation_score_DF, aes(x = TimePoint, y = differentiation_score_pca)) +
  geom_point(color = 'darkblue', size = 5 , alpha = 0.8) +
  ggtitle('Differentiation Score Across Timepoints
  via Weighted Sum (PCA 1)') +
  theme_minimal(base_size = 15) +
  labs(x = 'Timepoints', y = 'Differentiation Score') +
  theme(plot.title = element_text(hjust = 0.5, size = 17))

a
```


```{r}
pdf("figures/differentiationScore_plot.pdf",
    height = 5,
    width = 8)

a

dev.off()

```

# Apply differentiation score to cell data
```{r}
all(rownames(pheno_cells) == colnames(cts_cells))
dds_cells <- DESeqDataSetFromMatrix(countData = cts_cells, colData = pheno_cells, design = ~ Sex)

x <- dim(dds_cells)

keep <- rowMeans(counts(dds_cells)) >= 10
dds_cells <- dds_cells[keep,]

dds_cells <- DESeq(dds_cells)
normalized_counts_cells <- as.matrix(assay(vst(dds_cells, blind = T)))

# Extract gene names (rownames) from differentiation dataset
new_data <- normalized_counts_cells[rownames(normalized_counts_cells) %in% names(pc1_loadings), ]

pc1_loadings2 <- pc1_loadings[names(pc1_loadings) %in% rownames(new_data)]

new_data <- new_data[names(pc1_loadings2), ]
identical(names(pc1_loadings2), rownames(new_data))

# Apply previous PC1 loadings as weights to the new normalized data
cell_weighted_sum <- apply(new_data, 2, function(sample) {
  sum(sample * pc1_loadings2, na.rm = TRUE)
})

# Min-max scaling
min_score_new <- min(cell_weighted_sum)
max_score_new <- max(cell_weighted_sum)
cell_differentiation_score <- (cell_weighted_sum - min_score) / (max_score - min_score)

# ensures that any scores below 0 are capped at 0 and any scores above 1 are capped at 1
#cell_differentiation_score <- pmin(pmax(cell_differentiation_score, 0), 1)
cell_differentiation_scoreDF <- as.data.frame(cell_differentiation_score)
cell_differentiation_scoreDF

write.csv(cell_differentiation_scoreDF, "data/cell_differentiation_scoreDF.csv")
```

# Table 2: Classify samples according to differentiation score
```{r}
scores <- differentiation_score_DF$differentiation_score_pca
cell_classified_scoreDF2 <- cell_differentiation_scoreDF %>% 
  rownames_to_column(var = "ID") %>% 
  mutate(Diff_timePoint = case_when(
    cell_differentiation_score <= scores[1] ~ "T1",
    cell_differentiation_score > scores[1] & cell_differentiation_score <= scores[2] ~ "T2",
    cell_differentiation_score > scores[2] & cell_differentiation_score <= scores[3] ~ "T3",
    cell_differentiation_score > scores[3] & cell_differentiation_score <= scores[4] ~ "T4",
    cell_differentiation_score > scores[4] & cell_differentiation_score <= scores[5] ~ "T5",
    cell_differentiation_score > scores[5] & cell_differentiation_score <= 1.0 ~ "T6",
    cell_differentiation_score > 1.0 ~ "T6_plus",
    TRUE ~ NA_character_  # In case some values don't fit the thresholds
  ))

#saveRDS(cell_classified_scoreDF2, "rds/cell_classified_scoreDF2.rds")
write.csv(cell_classified_scoreDF2, "data/cell_classified_scoreDF2.csv")
```


# Session information
```{r}
sessionInfo()   
``` 

